Hypertension and microbiome

### 2019-06-07 articletwo.Rmd
First full run on centrifuge

options(max.print=1000)

@@

    saveRDS(maaslin, file = "rds/maaslin.rds")
} else {
    maaslin <- readRDS(file = "rds/maaslin.rds")
}

@@

alphadiversity <- list(
    min = calculateglm(dset.alpha,
                       modelstring = sprintf("%%s ~ %s + diversity_shannon",
                                             paste0(var.CL.min, collapse = " + ")),
                       filterstr = "shannon"),
    max = calculateglm(dset.alpha,
                       modelstring = sprintf("%%s ~ %s + diversity_shannon",
                                             paste0(var.CL, collapse = " + ")),
                       filterstr = "shannon"))
```

@@

```{r pcoa plot}
pcoa.vectors <- pcoa.ordinate$vectors %>% as.data.frame
eig.fracs <- pcoa.ordinate$values$Relative_eig
pcoa.df <- merge(pcoa.vectors %>% select(Axis.1, Axis.2, Axis.3),
                 meta(pseq.species), by=0, all = TRUE) %>%
    mutate(STATETREATMENT = factor(paste0(HYPERTENSION, ANYDRUG),
                                   levels = c("11", "10", "01", "00")),
           gMAP = cut(oMAP, seq(60, 160, 15)))

pcoa.plot <- ggplot(data=pcoa.df, aes(x=Axis.1, y=Axis.3, color = HYPERTENSION)) +
    geom_point(size=0.1, alpha=0.4) +
    xlab(sprintf("PCoA Axis 1 (%.1f%%)", 100*eig.fracs[1])) +
    ylab(sprintf("PCoA Axis 3 (%.1f%%)", 100*eig.fracs[3])) +
    scale_colour_manual(name = "State and treatment",
                        labels = c("Normotensive", "Hypertensive"),
                        breaks=c(0, 1),
                        values = c("blue", "red")) +   
    coord_fixed() +
    scale_x_continuous(breaks=seq(-1, 1, .25)) +
    scale_y_continuous(breaks=seq(-1, 1, .25)) +
    theme_classic() +
    guides(colour = guide_legend(override.aes = list(size=4))) +
    theme(text = element_text(size=14),
          legend.title = element_blank(),
          legend.text=element_text(size=10),
          legend.justification=c(1,0),
          legend.position=c(1,0),
          legend.margin=margin(t = 1, unit='mm'),
          legend.background = element_rect(fill=alpha('gray', 0.1)))

ggsave(file = "cache/pcoa-species.png", plot=pcoa.plot, height=4, width=5, units = "in", dpi = 300)
```

@@

Genera with top R2 with PCoA axes

```{r top generas}
abund.genus <- abundances(pseq.genus, transform = "compositional") %>%
    t %>%
    as.data.frame %>%
    tibble::rownames_to_column("id") %>%
    setNames(tolower(replace.brackets(names(.))))

@@

```{r maaslin}
maas.results.df.signf <- maaslin$genus %>%
    merge(names.dset %>% select(Covariate, Category, Name), by.x ="Variable", by.y = "Covariate") %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
                  Feature = gsub("_Bacteria", "", Feature),
                  beta = Coefficient) %>%
    dplyr::filter(Qval < 0.05) %>%
    arrange(Feature)

maas.results.df.signf %>% pull(Feature) %>% sort %>% unique %>% paste(., collapse=", ")

@@

maaslinplot <- ggplot(maas.tables.df, aes(x = Feature,
                                          y = reorder(Name, my.plot.order(Name)),
                                          fill = p.groups)) +
    geom_tile(color = "white", aes(fill = "NA")) +
    geom_tile(color = "white") +
    scale_fill_manual(values = mycolors, name = "P-value", na.translate = F) +
    scale_x_discrete() +
    coord_fixed() +
    xlab("") +
    ylab("") +
    theme_classic(10) +  
    theme(axis.text.x = element_text(angle = 90, hjust = 1.0, vjust=0.5, size=10),
          axis.text.y = element_text(size=10))
ggsave(file = "cache/maaslin.png", plot = maaslinplot, height = 3, width = 6, dpi = 300)
```

@@

```{r scatter plots}
scatter.signf.bacteria <- paste0(maas.results.df.signf %>% pull(Feature) %>% unique %>% tolower, ".bacteria")

@@

scatter.abund <- abundances(pseq.genus, transform = "clr") %>%
    t %>%
    as.data.frame %>%
    tibble::rownames_to_column("id") %>%
    setNames(tolower(replace.brackets(names(.)))) %>%
    select(id, scatter.signf.bacteria) %>%
    mutate_at(.vars = vars(scatter.signf.bacteria), .funs = funs(myscale(.)))

@@

    feature <- firstup(gsub(".bacteria", "", replace.brackets(abund)))
    qval.maaslin <- maas.results %>%
        dplyr::filter(Feature == feature, Value == rr) %>%
        nrow

@@

```{r scatter plot at genus level for BP, fig.width=12, fig.height=20}
scatterplot <- do.call("grid.arrange", c(scatterplots, ncol=4))
ggsave(file = "cache/scatter.png", plot = scatterplot, height = 20, width = 10, dpi = 300)
```

@@

```{r check odds model assumptions}
g.odds.autoplot <- autoplot(glm.hypertension.lactogroup,
                            which = 1:4,
                            nrow = 1,
                            ncol = 2,
                            label.size = 3,
                            size = 0.1) +
    theme_minimal()
ggsave(file = "cache/odds-autoplot.png", plot=g.odds.autoplot, height=10, width=6, dpi = 300)
```

@@

```{r forest plot lacto}
gsalt <- grid.arrange(
    myforestplot(adjmean.lacto.saltgroup,
                 xlab = "Dietary salt",
                 ylab = "Lactobacillus abundance",
                 subtitle = "A.",
                 my_y_scale = scale_y_continuous(limits = c(2.5,3), breaks = seq(2.5, 3.0, 0.1))),
    myforestplot(odds.hypertension.lactogroup,
                 xlab = "Lactobacillus abundance",
                 ylab = "Odds of hypertension",
                 subtitle = "B.",
                 my_y_scale = scale_y_log10(limits = c(0.5, 2), breaks = seq(0.5, 2.0, 0.5)),
                 interceptone = TRUE),
    ncol=2)

>>>>

options(max.print=100)

@@

    saveRDS(maaslin, file = "rds/maaslin.rds")
} else {
    maaslin <- readRDS(file = "rds/maaslin.rds")
}
```

@@

alphadiversity <- list(
    min = calculateglm(dset.alpha,
                       modelstring = sprintf("%%s ~ %s + diversity_shannon",
                                             paste0(var.CL.min, collapse = " + ")),
                       filterstr = "shannon"),
    max = calculateglm(dset.alpha,
                       modelstring = sprintf("%%s ~ %s + diversity_shannon",
                                             paste0(var.CL, collapse = " + ")),
                       filterstr = "shannon"))

diversities <- mergediversities(alphadiversity, adonis.species)
```

@@

```{r pcoa plot}
pcoa.vectors <- pcoa.ordinate$vectors %>% as.data.frame
eig.fracs <- pcoa.ordinate$values$Relative_eig
pcoa.df <- merge(pcoa.vectors %>% select(Axis.1, Axis.2, Axis.3),
                 meta(pseq.species), by=0, all = TRUE) %>%
    mutate(STATETREATMENT = factor(paste0(HYPERTENSION, ANYDRUG),
                                   levels = c("11", "10", "01", "00")),
           gMAP = cut(oMAP, seq(60, 160, 15))) %>%
    gather(key, Axis, Axis.2, Axis.3)

axis_labeller <- function(variable, value){
    axis.labels <- list("Axis.2" = sprintf("A. PCoA Axis 2 (%.1f%%)", 100*eig.fracs[2]),
                        "Axis.3" = sprintf("B. PCoA Axis 3 (%.1f%%)", 100*eig.fracs[3]))
    return(axis.labels[value])
}

pcoa.plot <- ggplot(data=pcoa.df, aes(x=Axis.1, y=Axis, color = HYPERTENSION)) +
    facet_wrap(~key, labeller = axis_labeller) +
    geom_point(size=0.1, alpha=0.4) +
    xlab(sprintf("PCoA Axis 1 (%.1f%%)", 100*eig.fracs[1])) +
    ylab(sprintf("Subsequent PCoA Axes", 100*eig.fracs[3])) +
    scale_colour_manual(name = "State and treatment",
                        labels = c("Normotensive", "Hypertensive"),
                        breaks=c(0, 1),
                        values = c("blue", "red")) +   
    coord_fixed() +
    scale_x_continuous(breaks=seq(-1, 1, .25)) +
    scale_y_continuous(breaks=seq(-1, 1, .25)) +
    theme_classic() +
    guides(colour = guide_legend(override.aes = list(size=4))) +
    theme(text = element_text(size=14),
          legend.position = "none",
          strip.background = element_blank(),
          strip.placement = "outside")

ggsave(file = "cache/pcoa-species.png", plot=pcoa.plot, height=4, width=6, units = "in", dpi = 300)
```

@@

```{r top generas, eval = FALSE, include = FLASE}
abund.genus <- abundances(pseq.genus, transform = "compositional") %>%
    t %>%
    as.data.frame %>%
    tibble::rownames_to_column("id") %>%
    setNames(tolower(replace.brackets(names(.))))

@@

```{r maaslin}
maas.results.df.signf <- maaslin %>%
    merge(names.dset %>% select(Covariate, Category, Name), by.x ="Variable", by.y = "Covariate") %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
                  Feature = gsub("_Bacteria", "", Feature),
                  beta = Coefficient) %>%
    dplyr::filter(Qval < 0.05) %>%
    arrange(Feature)

@@

maas.tables.df %>% pull(beta)%>% summary

maaslinplot <- ggplot(maas.tables.df, aes(x = Feature, y = reorder(Name, my.plot.order(Name)), fill = beta)) +
    geom_tile(aes(fill = 1)) +
    geom_tile(colour="white", size=0.25) +
    scale_fill_distiller(palette = "RdBu", name = "Effect size",
                         limits=c(-0.003, 0.003), breaks = c(-0.003, 0, 0.003), na.value="grey30") +
    coord_fixed() +
    xlab("") +
    ylab("") +
    theme_classic(10) +  
    theme(axis.text.x = element_text(angle = 90, hjust = 1.0, vjust=0.5, size=10),
          axis.text.y = element_text(size=10))

ggsave(file = "cache/maaslin.png", plot = maaslinplot, height = 3, width = 6, dpi = 300)
```

@@

```{r scatter plots}
scatter.signf.bacteria <- maas.results.df.signf %>% pull(Feature) %>% unique

@@

scatter.abund <- abundances(pseq.genus, transform = "clr") %>%
    t %>%
    as.data.frame %>%
    tibble::rownames_to_column("id") %>%
    select(id, scatter.signf.bacteria) %>%
    mutate_at(.vars = vars(scatter.signf.bacteria), .funs = funs(myscale(.)))

@@

    feature <- replace.brackets(abund)

    qval.maaslin <- maas.results %>%
        dplyr::filter(Feature == feature, Value == rr) %>%
        nrow

@@

```{r scatter plot at genus level for BP, fig.width=12, fig.height=10}
scatterplot <- do.call("grid.arrange", c(scatterplots, ncol=4))
ggsave(file = "cache/scatter.png", plot = scatterplot, height = 10, width = 10, dpi = 300)
```

@@

@@

```{r forest plot lacto}
odds.hypertension.lactogroup
gsalt <- grid.arrange(
    myforestplot(adjmean.lacto.saltgroup,
                 xlab = "Dietary salt",
                 ylab = "Lactobacillus abundance",
                 subtitle = "A.",
                 my_y_scale = scale_y_continuous(limits = c(4.1,4.7), breaks = seq(4, 5, 0.1))),
    myforestplot(odds.hypertension.lactogroup,
                 xlab = "Lactobacillus abundance",
                 ylab = "Odds of hypertension",
                 subtitle = "B.",
                 my_y_scale = scale_y_log10(limits = c(0.4, 1.5), breaks = seq(0.5, 2.0, 0.5)),
                 interceptone = TRUE),
    ncol=2)
